Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2018-06-14 submission/manuscript.tex
add intro fixes suggested by Nikc

Colorectal cancer (CRC) is currently the third leading cancer-related
cause of death within the US (1, 2). Although there is a genetic
component to the disease, the environment has been attributed as an
important and large risk factor for CRC (3). This has lead to the
investigation of how the microbiota may exacerbate or cause tumorigensis
(4--6) and whether the bacterial community is altered (7, 8). Multiple
reports in case/control studies on the microbiota and CRC have
identified taxa commonly associated with the normal colons to be
decreased in patients with carcinoma tumors (9--11). Many of these taxa
actively produce short-chain fatty acids (SCFAs) from fiber that are a
part of our general diet (12). The most extensively studied of these
SCFAs are acetate, butyrate, and propionate (13). Overall, CRC is a
growing health problem and the microbiota through SCFAs have shown
promise as a potential treatment option.

Specific SCFAs, like butyrate, have shown positive results for CRC
treatment within model systems (14). Butyrate has been shown to inhibit
cancer cell growth in \emph{in vitro} systems (15). Additionally,
supplementation with food sources that bacteria use to create these
SCFAs may also be able to confer beneficial effects. For example, fiber
supplementation in mouse models of CRC caused an overall reduction in
tumor burden while also increasing SCFA concentrations (16). Although
these model systems provide important preliminary evidence towards the
ability of SCFAs to reduce and treat tumors, the studies reporting
benefit in humans has been less convincing.

There is a lack of evidence on the benefit of increasing SCFA
concentrations to protect against CRC in human populations. The initial
case/control studies that investigated SCFA concentrations in CRC found
that patients with carcinomas had lower concentrations of acetate,
butyrate, and propionate versus either patients with adenomas or
individuals with normal colons (17). Although this would argue that
increasing SCFA concentrations could be protective against
tumorigenesis, fiber supplementation in randomized controlled trials
have consistently failed to protect against tumor recurrence (18). These
findings would argue against the utility of treatments that aim to use
SCFAs to reduce or protect against tumorigenesis. Given the lack of
clear evidence in human studies of the benefit of SCFAs in CRC, there is
a need for more investigation into this area.

Our current study will fill some of the current gaps in the literature
that relate to the study of SCFAs and CRC in human populations.
Specifically, it will attempt to replicate previous case/control
findings on SCFA concentrations in individuals with and without tumors.
Additionally, by increasing the adenoma tumor cohort it we will identify
if the observation of a continuous gradual reduction in SCFA
concentrations occurs as tumor severity increases. It also will build
upon these observations and assess the utility of using SCFAs and
Operational Taxonomic Units (OTUs) as a potential risk stratification
tool of colorectal tumors (adenoma or carcinoma). This study will
provide important information on the replicability of previous findings
in humans by extensively studying how SCFAs could be associated with
colorectal tumors.

To accomplish this task we will directly measure the concentration of
acetate, butyrate, and propionate within fecal samples for two different
groups. The first group had a sample obtained at a single cross
sectional point in time while the second group had samples obtained
before (pre-) and after (post-)treatment for colorectal tumors.
Additionally, we also assessed whether adding SCFA concentrations to OTU
data improves the classification of adenoma or carcinoma using the
Random Forest algorithm (19), used PICRUSt (20) and metagenomic
sequencing to assess the genes involved in SCFA synthesis, and analyzed
how well 16S rRNA gene sequencing can predict SCFA concentrations. This
investigation provides additional information as to whether the taxa
that are decreased in carcinoma use SCFAs to help protect against colon
tumorigenesis.

@@

Much of the previous research on SCFA benefit to human CRC has been
illustrated in model systems (14). Many SCFAs are produced through the
breakdown of fiber (12) and a recent study in mice found that fiber
supplementation increased SCFA concentrations and decreased tumor burden
(16). Additionally, SCFAs such as butyrate can inhibit tumor growth in
\emph{in vitro} experiments (15). Yet, observations in humans has been
mixed. Previous case/control studies found associations with lower SCFA
concentrations in individuals with carcinoma tumors (17). However,
individual randomized controlled trials and a recent meta-analysis on
fiber supplementation to prevent tumor recurrence has found no benefit
(18, 21). Our results align with what has been reported in
randomized-controlled trials, that SCFAs do not provide general
protection against colorectal tumors. It is possible though that there
are specific instances where SCFAs may be beneficial.

One limitation of ours and others research into the effect of SCFAs in
CRC has been that all tumors are treated as the same type. However,
there are known differences in the types of mutations that occur (22)
and treating all tumors as equal may actually hide any benefit that
could be found in specific subsets of individuals. Similar to the idea
of specific immunotherapy for specific tumors (23), SCFAs may have
beneficial effects for specific types of colorectal tumors. Future
research will need to test if this is a valid hypothesis. Although there
are limitations with the current research on SCFAs and colorectal
tumors, evidence is building that these specific metabolites may not be
protective. Yet, taxa that are associated with SCFA production are
consistently higher in normal colons than those with carcinomas (7, 8,
24).

The potential protection against colorectal cancers may not be from
SCFAs even though taxa associated with their production are higher in
individuals with normal colons versus patients with carcinomas (7, 8,
24). Protection could be via a different pathway and by extension other
metabolites that have not been extensively studied. Alternatively,
protection may not occur via a metabolite but instead through niche
exclusion of mouth-associated microbes (e.g. \emph{Fusobacterium},
\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (9,
10)). The idea of niche exclusion is similar to how the community
protects against \emph{Clostridium difficile} infection (25) with
chronic inflammation replacing the role of antibiotics. Although we did
not find lower concentrations of SCFAs associated with colorectal
tumors, we think that there are many exciting new avenues to explore
because of these results

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (26, 27). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment of a patient's tumor
(n=67). For patients undergoing treatment for their tumor the length of
time between their initial and follow up sample ranged from 188 - 546
days. Our use of treatment has been previously defined as encompassing
removal of a tumor with or without chemotherapy and radiation (26).
Diagnosis of tumor was made by colonoscopic examination and
histopathological review of biopsies obtained (26, 27). The University
of Michigan Institutional Review Board approved the study and informed
consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Declaration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol that used a High-Performance Liquid Chromatography (HPLC)
machine (28). The following changes to this protocol included the use of
frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the actual
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These methodological changes did not
affect the overall median concentrations of these SCFAs between the two
studies (see Table 1 (28) and Figure 1 here).

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
previously described (26, 29, 30). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity using the OptClust algorithm (31).

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (20) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formatted table using the biom convert function, the subsequent biom
file was processed with the `normalize\_by\_copy\_number.py' function,
and subsequent imputed metagenomes created using the
`predict\_metagenomes.py' function.

\textbf{Obtaining Operational Protein Families from metagenomes.} A
subset of the cross-sectional group (n=490) containing a total of 85
individuals (normal n=29, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq using 125 bp paired end reads and a
previously described method (32). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (33). Open Reading Frames (ORFs) were
identified using Prodigal (34), counts generated using Diamond (35),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (36), and OPF alignment used the KEGG database (37).

\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
located near the end of the pathways involved in the synthesis of
acetate, butyrate, and propionate were analyzed for any differences
between individuals with normal colons and those with tumors. These
genes were based on pathways from KEGG as well as previous research (37,
38) and a list can be found in the supplemental material {[}Table S1{]}.

\textbf{Random Forest models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for classification models (19). This was
repeated on 100 different 80/20 splits of the data to generate a
reasonable range for the AUC of the model. The reported AUCs, unless
otherwise specified, are for the test sets. The classification models
were built to group normal versus adenoma, normal versus carcinoma, and
high versus low SCFA concentrations.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (39). Generally, a Kruskal-Walis rank sum
test with a Dunn's post-hoc test was used to assess differences between
the groups used. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (40). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, we assessed whether OTUs
alone or OTUs and SCFAs were better able to classify individuals with
and without tumor using Random Forest models. Next, the imputed gene
counts of important mediators of SCFA synthesis was tested.
Additionally, the counts generated for OPFs that matched important genes
involved with SCFA creation were analyzed. Finally, models to classify
high or low SCFA concentration based on the median of each SCFA and 16S
rRNA gene sequencing data was created using the Random Forest algorithm.

@@

\hypertarget{ref-Zackular2013}{}
4. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Iverson KD},
\textbf{Sadler WD}, \textbf{Petrosino JF}, \textbf{Chen GY},
\textbf{Schloss PD}. 2013. The gut microbiome modulates colon
tumorigenesis. mBio \textbf{4}:e00692--13--e00692--13.
doi:\href{https://doi.org/10.1128/mbio.00692-13}{10.1128/mbio.00692-13}.

\hypertarget{ref-Baxter2014}{}
5. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
colonization with human feces determines colonic tumor burden.
Microbiome \textbf{2}:20.
doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.

\hypertarget{ref-Zackular2015}{}
6. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Chen GY},
\textbf{Schloss PD}. 2015. Manipulation of the gut microbiota reveals
role in colon tumorigenesis. mSphere \textbf{1}:e00001--15.
doi:\href{https://doi.org/10.1128/msphere.00001-15}{10.1128/msphere.00001-15}.

\hypertarget{ref-Shah2017}{}
7. \textbf{Shah MS}, \textbf{DeSantis TZ}, \textbf{Weinmaier T},
\textbf{McMurdie PJ}, \textbf{Cope JL}, \textbf{Altrichter A},
\textbf{Yamal J-M}, \textbf{Hollister EB}. 2017. Leveraging
sequence-based faecal microbial community survey data to identify a
composite biomarker for colorectal cancer. Gut \textbf{67}:882--891.
doi:\href{https://doi.org/10.1136/gutjnl-2016-313189}{10.1136/gutjnl-2016-313189}.

\hypertarget{ref-meta_analysis_crc_Sze2018}{}
8. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors.
doi:\href{https://doi.org/10.1101/285486}{10.1101/285486}.

\hypertarget{ref-Zeller2014}{}
9. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
\textbf{Amiot A}, \textbf{Bohm J}, \textbf{Brunetti F},
\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Nhieu JTV},
\textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
for early-stage detection of colorectal cancer. Molecular Systems
Biology \textbf{10}:766--766.
doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.

\hypertarget{ref-Baxter2016}{}
10. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-Kostic2011}{}
11. \textbf{Kostic AD}, \textbf{Gevers D}, \textbf{Pedamallu CS},
\textbf{Michaud M}, \textbf{Duke F}, \textbf{Earl AM}, \textbf{Ojesina
AI}, \textbf{Jung J}, \textbf{Bass AJ}, \textbf{Tabernero J},
\textbf{Baselga J}, \textbf{Liu C}, \textbf{Shivdasani RA},
\textbf{Ogino S}, \textbf{Birren BW}, \textbf{Huttenhower C},
\textbf{Garrett WS}, \textbf{Meyerson M}. 2011. Genomic analysis
identifies association of fusobacterium with colorectal carcinoma.
Genome Research \textbf{22}:292--298.
doi:\href{https://doi.org/10.1101/gr.126573.111}{10.1101/gr.126573.111}.

\hypertarget{ref-Holscher2017}{}
12. \textbf{Holscher HD}. 2017. Dietary fiber and prebiotics and the
gastrointestinal microbiota. Gut Microbes \textbf{8}:172--184.
doi:\href{https://doi.org/10.1080/19490976.2017.1290756}{10.1080/19490976.2017.1290756}.

\hypertarget{ref-Louis2016}{}
13. \textbf{Louis P}, \textbf{Flint HJ}. 2016. Formation of propionate
and butyrate by the human colonic microbiota. Environmental Microbiology
\textbf{19}:29--41.
doi:\href{https://doi.org/10.1111/1462-2920.13589}{10.1111/1462-2920.13589}.

\hypertarget{ref-test_OKeefe2016}{}
14. \textbf{O'Keefe SJD}. 2016. Diet, microorganisms and their
metabolites and colon cancer. Nature Reviews Gastroenterology \&
Hepatology \textbf{13}:691--706.
doi:\href{https://doi.org/10.1038/nrgastro.2016.165}{10.1038/nrgastro.2016.165}.

\hypertarget{ref-Encarnao2018}{}
15. \textbf{Encarnação JC}, \textbf{Pires AS}, \textbf{Amaral RA},
\textbf{Gonçalves TJ}, \textbf{Laranjo M}, \textbf{Casalta-Lopes JE},
\textbf{Gonçalves AC}, \textbf{Sarmento-Ribeiro AB}, \textbf{Abrantes
AM}, \textbf{Botelho MF}. 2018. Butyrate, a dietary fiber derivative
that improves irinotecan effect in colon cancer cells. The Journal of
Nutritional Biochemistry \textbf{56}:183--192.
doi:\href{https://doi.org/10.1016/j.jnutbio.2018.02.018}{10.1016/j.jnutbio.2018.02.018}.

\hypertarget{ref-Bishehsari2018}{}
16. \textbf{Bishehsari F}, \textbf{Engen P}, \textbf{Preite N},
\textbf{Tuncil Y}, \textbf{Naqib A}, \textbf{Shaikh M}, \textbf{Rossi
M}, \textbf{Wilber S}, \textbf{Green S}, \textbf{Hamaker B},
\textbf{Khazaie K}, \textbf{Voigt R}, \textbf{Forsyth C},
\textbf{Keshavarzian A}. 2018. Dietary fiber treatment corrects the
composition of gut microbiota, promotes SCFA production, and suppresses
colon carcinogenesis. Genes \textbf{9}:102.
doi:\href{https://doi.org/10.3390/genes9020102}{10.3390/genes9020102}.

\hypertarget{ref-Ohigashi2013}{}
17. \textbf{Ohigashi S}, \textbf{Sudo K}, \textbf{Kobayashi D},
\textbf{Takahashi O}, \textbf{Takahashi T}, \textbf{Asahara T},
\textbf{Nomoto K}, \textbf{Onodera H}. 2013. Changes of the intestinal
microbiota, short chain fatty acids, and fecal pH in patients with
colorectal cancer. Digestive Diseases and Sciences
\textbf{58}:1717--1726.
doi:\href{https://doi.org/10.1007/s10620-012-2526-4}{10.1007/s10620-012-2526-4}.

\hypertarget{ref-Yao2017}{}
18. \textbf{Yao Y}, \textbf{Suo T}, \textbf{Andersson R}, \textbf{Cao
Y}, \textbf{Wang C}, \textbf{Lu J}, \textbf{Chui E}. 2017. Dietary fibre
for the prevention of recurrent colorectal adenomas and carcinomas.
Cochrane Database of Systematic Reviews.
doi:\href{https://doi.org/10.1002/14651858.cd003430.pub2}{10.1002/14651858.cd003430.pub2}.

>>>>

Colorectal cancer (CRC) is currently the third highest cancer-related
cause of death within the US (1, 2). Although there is a genetic
component to the disease, the environment is attributed to being a
larger risk factor for CRC (3). These environmental risk factors include
but are not limited to smoking cigarettes, diet, and the microbiota
(4--6). Many of these environmental risk factors are capable of being
modified, and this has lead to the investigation of how the microbiota
may exacerbate or cause tumorigensis (7--9) and whether the bacterial
community is altered (10, 11). Multiple reports in case/control studies
have identified bacterial taxa commonly associated with individuals
without tumors to be decreased in patients with carcinoma tumors
(11--13). Many of these taxa within individuals wihtout tumors actively
produce short-chain fatty acids (SCFAs) from fiber that are a part of
our general diet (14). The most extensively studied of these SCFAs are
acetate, butyrate, and propionate (15). Overall, the specific bacterial
taxa of the microbiota that create SCFAs are an attractive target to
modulate the risk of CRC.

Specific SCFAs, like butyrate, have shown positive results for CRC
treatment within model systems (16). Butyrate has been shown to inhibit
cancer cell growth in \emph{in vitro} systems (17). Additionally,
supplementation with food sources that bacteria use to create these
SCFAs may also be able to confer beneficial effects. For example, fiber
supplementation in mouse models of CRC caused an overall reduction in
tumor burden while also increasing SCFA concentrations (18). Although
these model systems provide important preliminary evidence towards the
ability of SCFAs to reduce and treat tumors, the studies reporting
benefit in humans has been less convincing.

There is a lack of evidence on the benefit of increasing SCFA
concentrations to protect against CRC in human populations. The initial
case/control studies that investigated SCFA concentrations in CRC found
that patients with carcinomas had lower concentrations of acetate,
butyrate, and propionate versus either patients with adenomas or
individuals without colon tumors (19). Although this would argue that
increasing SCFA concentrations could be protective against
tumorigenesis, fiber supplementation in randomized controlled trials
have consistently failed to protect against tumor recurrence (20). These
findings would argue against the utility of treatments that aim to use
SCFAs to reduce or protect against tumorigenesis. Given the lack of
clear evidence in human studies of the benefit of SCFAs in CRC, there is
a need for more investigation into this area.

Our study fills some of the current gaps in the literature that relate
to the study of SCFAs and CRC in human populations. Specifically, it
tests previous case/control findings on SCFA concentrations in
individuals with and without tumors. We also test previous suggestions
that there is a continuous reduction in SCFA concentrations as tumor
severity increases by increasing the number of patients with adenomas in
our study. Additionally, we build upon these observations and assesses
the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a
risk stratification tool of colorectal tumors (adenoma or carcinoma).
Collectively, this study provides important information on the
replicability of previous findings in humans by extensively studying how
SCFAs are associated with colorectal tumors.

To accomplish this task we directly measured the concentration of
acetate, butyrate, and propionate within fecal samples for two different
groups. The first group had a sample obtained at a single cross
sectional point in time while the second group had samples obtained
before (pre-) and after (post-)treatment for colorectal tumors.
Additionally, we (i) assessed the affect adding SCFA concentrations to
OTU data had on classification of patients with adenoma or carcinoma
using the Random Forest algorithm (21), (ii) used PICRUSt (22) and
metagenomic sequencing to assess the presence of genes involved in SCFA
synthesis, and (iii) analyzed how well 16S rRNA gene sequencing predicts
SCFA concentrations. This investigation provides additional information
as to whether SCFAs are decreased in patients with colorectal tumors and
provides context as to whether targeting taxa to increase SCFA
concentrations is a viable option to protect against colon
tumorigenesis.

@@

Much of the previous research on SCFA benefit to human CRC has been
illustrated in model systems (16). Many SCFAs are produced through the
breakdown of fiber (14) and a recent study in mice found that fiber
supplementation increased SCFA concentrations and decreased tumor burden
(18). Additionally, SCFAs such as butyrate can inhibit tumor growth in
\emph{in vitro} experiments (17). Yet, observations in humans has been
mixed. Previous case/control studies found associations with lower SCFA
concentrations in individuals with carcinoma tumors (19). However,
individual randomized controlled trials and a recent meta-analysis on
fiber supplementation to prevent tumor recurrence has found no benefit
(20, 23). Our results align with what has been reported in
randomized-controlled trials, that SCFAs do not provide general
protection against colorectal tumors. It is possible though that there
are specific instances where SCFAs may be beneficial.

One limitation of ours and others research into the effect of SCFAs in
CRC has been that all tumors are treated as the same type. However,
there are known differences in the types of mutations that occur (24)
and treating all tumors as equal may actually hide any benefit that
could be found in specific subsets of individuals. Similar to the idea
of specific immunotherapy for specific tumors (25), SCFAs may have
beneficial effects for specific types of colorectal tumors. Future
research will need to test if this is a valid hypothesis. Although there
are limitations with the current research on SCFAs and colorectal
tumors, evidence is building that these specific metabolites may not be
protective. Yet, taxa that are associated with SCFA production are
consistently higher in normal colons than those with carcinomas (10, 11,
26).

The potential protection against colorectal cancers may not be from
SCFAs even though taxa associated with their production are higher in
individuals with normal colons versus patients with carcinomas (10, 11,
26). Protection could be via a different pathway and by extension other
metabolites that have not been extensively studied. Alternatively,
protection may not occur via a metabolite but instead through niche
exclusion of mouth-associated microbes (e.g. \emph{Fusobacterium},
\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (12,
13)). The idea of niche exclusion is similar to how the community
protects against \emph{Clostridium difficile} infection (27) with
chronic inflammation replacing the role of antibiotics. Although we did
not find lower concentrations of SCFAs associated with colorectal
tumors, we think that there are many exciting new avenues to explore
because of these results

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (28, 29). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment of a patient's tumor
(n=67). For patients undergoing treatment for their tumor the length of
time between their initial and follow up sample ranged from 188 - 546
days. Our use of treatment has been previously defined as encompassing
removal of a tumor with or without chemotherapy and radiation (28).
Diagnosis of tumor was made by colonoscopic examination and
histopathological review of biopsies obtained (28, 29). The University
of Michigan Institutional Review Board approved the study and informed
consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Declaration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol that used a High-Performance Liquid Chromatography (HPLC)
machine (30). The following changes to this protocol included the use of
frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the actual
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These methodological changes did not
affect the overall median concentrations of these SCFAs between the two
studies (see Table 1 (30) and Figure 1 here).

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
previously described (28, 31, 32). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity using the OptClust algorithm (33).

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (22) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formatted table using the biom convert function, the subsequent biom
file was processed with the `normalize\_by\_copy\_number.py' function,
and subsequent imputed metagenomes created using the
`predict\_metagenomes.py' function.

\textbf{Obtaining Operational Protein Families from metagenomes.} A
subset of the cross-sectional group (n=490) containing a total of 85
individuals (normal n=29, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq using 125 bp paired end reads and a
previously described method (34). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (35). Open Reading Frames (ORFs) were
identified using Prodigal (36), counts generated using Diamond (37),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (38), and OPF alignment used the KEGG database (39).

\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
located near the end of the pathways involved in the synthesis of
acetate, butyrate, and propionate were analyzed for any differences
between individuals with normal colons and those with tumors. These
genes were based on pathways from KEGG as well as previous research (39,
40) and a list can be found in the supplemental material {[}Table S1{]}.

\textbf{Random Forest models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for classification models (21). This was
repeated on 100 different 80/20 splits of the data to generate a
reasonable range for the AUC of the model. The reported AUCs, unless
otherwise specified, are for the test sets. The classification models
were built to group normal versus adenoma, normal versus carcinoma, and
high versus low SCFA concentrations.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (41). Generally, a Kruskal-Walis rank sum
test with a Dunn's post-hoc test was used to assess differences between
the groups used. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (42). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, we assessed whether OTUs
alone or OTUs and SCFAs were better able to classify individuals with
and without tumor using Random Forest models. Next, the imputed gene
counts of important mediators of SCFA synthesis was tested.
Additionally, the counts generated for OPFs that matched important genes
involved with SCFA creation were analyzed. Finally, models to classify
high or low SCFA concentration based on the median of each SCFA and 16S
rRNA gene sequencing data was created using the Random Forest algorithm.

@@

\hypertarget{ref-FlissIsakov2017}{}
4. \textbf{Fliss-Isakov N}, \textbf{Zelber-Sagi S}, \textbf{Webb M},
\textbf{Halpern Z}, \textbf{Kariv R}. 2017. Smoking habits are strongly
associated with colorectal polyps in a population-based case-control
study. Journal of Clinical Gastroenterology 1.
doi:\href{https://doi.org/10.1097/mcg.0000000000000935}{10.1097/mcg.0000000000000935}.

\hypertarget{ref-Lee2015}{}
5. \textbf{Lee J}, \textbf{Jeon JY}, \textbf{Meyerhardt JA}. 2015. Diet
and lifestyle in survivors of colorectal cancer. Hematology/Oncology
Clinics of North America \textbf{29}:1--27.
doi:\href{https://doi.org/10.1016/j.hoc.2014.09.005}{10.1016/j.hoc.2014.09.005}.

\hypertarget{ref-Kostic2011}{}
6. \textbf{Kostic AD}, \textbf{Gevers D}, \textbf{Pedamallu CS},
\textbf{Michaud M}, \textbf{Duke F}, \textbf{Earl AM}, \textbf{Ojesina
AI}, \textbf{Jung J}, \textbf{Bass AJ}, \textbf{Tabernero J},
\textbf{Baselga J}, \textbf{Liu C}, \textbf{Shivdasani RA},
\textbf{Ogino S}, \textbf{Birren BW}, \textbf{Huttenhower C},
\textbf{Garrett WS}, \textbf{Meyerson M}. 2011. Genomic analysis
identifies association of fusobacterium with colorectal carcinoma.
Genome Research \textbf{22}:292--298.
doi:\href{https://doi.org/10.1101/gr.126573.111}{10.1101/gr.126573.111}.

\hypertarget{ref-Zackular2013}{}
7. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Iverson KD},
\textbf{Sadler WD}, \textbf{Petrosino JF}, \textbf{Chen GY},
\textbf{Schloss PD}. 2013. The gut microbiome modulates colon
tumorigenesis. mBio \textbf{4}:e00692--13--e00692--13.
doi:\href{https://doi.org/10.1128/mbio.00692-13}{10.1128/mbio.00692-13}.

\hypertarget{ref-Baxter2014}{}
8. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
colonization with human feces determines colonic tumor burden.
Microbiome \textbf{2}:20.
doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.

\hypertarget{ref-Zackular2015}{}
9. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Chen GY},
\textbf{Schloss PD}. 2015. Manipulation of the gut microbiota reveals
role in colon tumorigenesis. mSphere \textbf{1}:e00001--15.
doi:\href{https://doi.org/10.1128/msphere.00001-15}{10.1128/msphere.00001-15}.

\hypertarget{ref-Shah2017}{}
10. \textbf{Shah MS}, \textbf{DeSantis TZ}, \textbf{Weinmaier T},
\textbf{McMurdie PJ}, \textbf{Cope JL}, \textbf{Altrichter A},
\textbf{Yamal J-M}, \textbf{Hollister EB}. 2017. Leveraging
sequence-based faecal microbial community survey data to identify a
composite biomarker for colorectal cancer. Gut \textbf{67}:882--891.
doi:\href{https://doi.org/10.1136/gutjnl-2016-313189}{10.1136/gutjnl-2016-313189}.

\hypertarget{ref-meta_analysis_crc_Sze2018}{}
11. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors.
doi:\href{https://doi.org/10.1101/285486}{10.1101/285486}.

\hypertarget{ref-Zeller2014}{}
12. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
\textbf{Amiot A}, \textbf{Bohm J}, \textbf{Brunetti F},
\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Nhieu JTV},
\textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
for early-stage detection of colorectal cancer. Molecular Systems
Biology \textbf{10}:766--766.
doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.

\hypertarget{ref-Baxter2016}{}
13. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-Holscher2017}{}
14. \textbf{Holscher HD}. 2017. Dietary fiber and prebiotics and the
gastrointestinal microbiota. Gut Microbes \textbf{8}:172--184.
doi:\href{https://doi.org/10.1080/19490976.2017.1290756}{10.1080/19490976.2017.1290756}.

\hypertarget{ref-Louis2016}{}
15. \textbf{Louis P}, \textbf{Flint HJ}. 2016. Formation of propionate
and butyrate by the human colonic microbiota. Environmental Microbiology
\textbf{19}:29--41.
doi:\href{https://doi.org/10.1111/1462-2920.13589}{10.1111/1462-2920.13589}.

\hypertarget{ref-test_OKeefe2016}{}
16. \textbf{O'Keefe SJD}. 2016. Diet, microorganisms and their
metabolites and colon cancer. Nature Reviews Gastroenterology \&
Hepatology \textbf{13}:691--706.
doi:\href{https://doi.org/10.1038/nrgastro.2016.165}{10.1038/nrgastro.2016.165}.

\hypertarget{ref-Encarnao2018}{}
17. \textbf{Encarnação JC}, \textbf{Pires AS}, \textbf{Amaral RA},
\textbf{Gonçalves TJ}, \textbf{Laranjo M}, \textbf{Casalta-Lopes JE},
\textbf{Gonçalves AC}, \textbf{Sarmento-Ribeiro AB}, \textbf{Abrantes
AM}, \textbf{Botelho MF}. 2018. Butyrate, a dietary fiber derivative
that improves irinotecan effect in colon cancer cells. The Journal of
Nutritional Biochemistry \textbf{56}:183--192.
doi:\href{https://doi.org/10.1016/j.jnutbio.2018.02.018}{10.1016/j.jnutbio.2018.02.018}.

\hypertarget{ref-Bishehsari2018}{}
18. \textbf{Bishehsari F}, \textbf{Engen P}, \textbf{Preite N},
\textbf{Tuncil Y}, \textbf{Naqib A}, \textbf{Shaikh M}, \textbf{Rossi
M}, \textbf{Wilber S}, \textbf{Green S}, \textbf{Hamaker B},
\textbf{Khazaie K}, \textbf{Voigt R}, \textbf{Forsyth C},
\textbf{Keshavarzian A}. 2018. Dietary fiber treatment corrects the
composition of gut microbiota, promotes SCFA production, and suppresses
colon carcinogenesis. Genes \textbf{9}:102.
doi:\href{https://doi.org/10.3390/genes9020102}{10.3390/genes9020102}.
